Seeing Double: Transient Dips and Photometric Binarity in PTFO$\,$8-8695

### 2020-04-22 paper/ms.tex
added words

\ptfo\ was initially identified as a member of the Orion$\,$OB1a
sub-association by \citet{briceno_cida_2005} through combined
photometry and spectroscopy.  Later work by \citet{briceno_25_2007}
clarified that \ptfo\ was in a kinematically distinct subgroup of
Orion~OB1a, named the ``25$\,$Ori'' group after its brightest
member. The 25$\,$Ori group had an isochrone age of 7--10$\,$Myr, and
had a smaller disk fraction than younger nearby sub-associations
\citep{hernandez_spitzer_2007}.

@@

\begin{figure*}[hbtp]
	\begin{center}
		\leavevmode
		\includegraphics[width=1\textwidth]{f7.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf \ptfo\ and its brethren.}
    Five transient and persistent flux dip stars selected based on
    their visual similarity to the short-period signal in \ptfo are as
    shown.  They include EPIC 204143627, EPIC 204321142, EPIC
    204270520, EPIC 205483258 (RIK-210), and EPIC 204787516.  RIK-210
    has the longest period of any of these objects.  EPIC 204787516
    has a less angular dip, and could also be an eclipsing binary.
    All stars have similar properties as \ptfo, but are in Upper Sco.
    We found these objects through studies by
    \citet{stauffer_orbiting_2017}, \citet{david_transient_2017}, and
    \citet{rebull_usco_2018}.
		\label{fig:brethren}
	}
\end{figure*}

When physical explanations are not immediately clear, taxonomy can be
a useful exercise.  To our knowledge, of order 10-20 lightcurves with
similar morphologies have been seen in surveys of low-mass weak-lined
T Tauri stars in regions including $\rho$ Oph, Upper Sco, Taurus, and perhaps
the Pleiades
\citep{rebull_rotation_2016,david_transient_2017,stauffer_orbiting_2017,stauffer_rotevol_2018,rebull_usco_2018,rebull_rotation_2020}.
These surveys were performed using K2 \citep{howell_k2_2014}.  We
downloaded a few of these lightcurves from MAST, opting for the
EVEREST reductions \citep{luger_everest_2016,luger_update_2018}.  They
are plotted in Figure~\ref{fig:brethren}.

These lightcurves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.
As \citet{stauffer_orbiting_2017} explain,
these  objects are morphologically distinct from ``scallop shell' lightcurves,
and are present in stars at more advanced evolutionary disk stages than
the ``dipper'' stars \citep{ansdell_young_2016,cody_manyfaceted_2018}.
The persistent and transient flux dip stars have the following properties.
\begin{enumerate}
  \item Weak-lined T Tauri stars.
  \item Spectral type typically M2-M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item Ages $\lesssim$ 100$\,$Myr.
  \item Stars typically lack spectroscopic accretion indicators 
  \item Phased lightcurves show shallow, angular dips, usually
    superposed on large-amplitude smooth variability. The latter
    is interpreted as stellar rotation.
  \item Rotation is usually between 0.5 to 2.0 days.
  \item Rarely any infrared excess (never any W4 detection; only a few
    cases with W3 detctions).
  \item Can show multiple dips per cycle ({\it e.g.}, EPIC 203692610,
    RIK-210).
  \item Dip depths, durations, and phases can vary over just a few cycles
    ({\it e.g.}, EPIC 204143627).
  \item Dip depths can change after flares.
  \item Rare at a population level: $\lesssim 1\%$ of relevant stars
    \citep{rebull_usco_2018}.
\end{enumerate}
The 10.74 hour signal in \ptfo\ meets essentially all of these
criteria.

A crucial additional point concerning the transient flux dips is
that their dip durations seem to scale linearly with
their periods \citep[][Figure~26]{stauffer_orbiting_2017}.
In the usual idealized limits, the transit duration of a point source
across the stellar disk scales as $T_0 \propto R_\star (P/M_\star)^{1/3}$
\citep{winn_exoplanet_2010}.
While the $\approx$0.5-day transient flux dip stars have dip
durations consistent with point sources, at longer periods of
1 to 5 days, the dip durations become many hours, and far
too long for a point-source origin.

@@

The evidence for binarity in \ptfo\ is as follows.  First, the star is
photometrically twice as bright as stars of the same color in its
kinematic group (Figure~\ref{fig:gaia}.  Second, it shows two distinct
photometric signals.  These two points alone are typically sufficient
to imply binarity \citep{stauffer_rotevol_2018}.  For the case of
\ptfo, there is a third line of evidence: the Gaia astrometry shows a
poor goodness-of-fit to a single-source model.  While this could be
caused by an additional companion or stellar variability, cluster
members that are equally if not more variable still show \ptfo\ to be
an outlier.  Therefore the astrometric excess is a suggestive third
line of evidence for binarity in \ptfo.  We take these three points
combined to imply that \ptfo\ is probably a binary.

% rumination: is the lack of an infrared excess even significant? In
% USco, we know late-M dwarfs show them. What about Orion? Distance is
% 4x, at least...
% Hernandez et al 2007a and 2007b show that there are some infrared
% excesses seen with Spitzer. But their completeness for fainter M
% dwarfs may be worse -- they see a disk fraction peaking at M0,
% instead of later as seen by e.g., Barenfeld+2016.
Based on the lack of an infrared excess seen by \citet{yu_tests_2015},
the primordial gas disks seem to be have been depleted around both
stars, and so the stars are no longer magnetically locked to their
disks.  This is consistent with the $\approx$half-day periodicities of
both signals: young disked M dwarfs typically rotate with periods of
two days or more due to magnetic locking \citep[{\it
e.g.},][]{rebull_rotation_2020}.  If the two stars are within
$\approx$50$\,$AU of each other, as required by the NIRC2 adaptive
optics imaging, then it would also be expected that the stars would
truncate the outer edges of their respective disks, in a manner seen
at the population level in exoplanetary systems
\citep{kraus_impact_2016,moe_impact_2019}.

The main physical question remaining is clearly what causes the
transient dipping. We have argued that this is an unsolved problem not
only for \ptfo\ but also for an emerging class of similar young
rapidly rotating M-dwarfs.

The explanation offered by \citet{stauffer_orbiting_2017} for this
variability class is that warm clouds of gas and dust orbit the star
at the Keplerian corotation radius.

Why???

What are the forces at play?

FIXME

Given the evidence for binarity, we believe that PTFO$\,$8-8695 is a
binary M dwarf in which one star shows the 11.96~hour rotation signal,
and the other shows a combination of transient dipping and stellar
rotation, similar to that observed in other young M dwarfs.  The
combined transient dipping and stellar rotation at the 10.74~hour
cycle was only detected with the TESS data.

The roughly half-day periods of both signals imply that

@@

Similarly, if the planet truly orbits every 10.74 hours, while the
star's equator spins every 11.96 hours, the situation is clearly
Darwin unstable.

  Given the available evidence,
  PTFO$\,$8-8695 seems consistent with the
  ``transient dipping'' phenomenology observed in many young M dwarfs.
  It seems rather unlikely to be a planet.

>>>>

\ptfo\ was initially identified as a member of the Orion$\,$OB1a
sub-association by \citet{briceno_cida_2005} through combined
photometry and spectroscopy.  Later work by \citet{briceno_25_2007}
clarified that \ptfo\ was in a kinematically distinct subgroup of
Orion~OB1a, named the ``25$\,$Ori'' group after its brightest
member. The 25$\,$Ori group had an isochrone age of 7--10$\,$Myr, and
had a smaller disk fraction than younger nearby sub-associations
\citep{hernandez_spitzer_ob1_2007}.

@@

\begin{figure*}[hbtp]
	\begin{center}
		\leavevmode
		\includegraphics[width=1\textwidth]{f7.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf \ptfo\ and its brethren.}
    Five transient and persistent flux dip stars selected based on
    their visual similarity to the short-period signal in \ptfo are as
    shown.  They include EPIC 204143627, EPIC 204321142, EPIC
    204270520, EPIC 205483258 (RIK-210), and EPIC 204787516.  RIK-210
    has the longest period of any of these objects.  EPIC 204787516
    has a less angular dip, and could also be an eclipsing binary.
    All stars have similar properties as \ptfo, but are in Upper~Sco.
    We found these objects through studies by
    \citet{stauffer_orbiting_2017}, \citet{david_transient_2017}, and
    \citet{rebull_usco_2018}.
		\label{fig:brethren}
	}
\end{figure*}

When physical explanations are not forthcoming, taxonomy can be a
useful exercise.  To our knowledge, of order 10-20 lightcurves with
similar morphologies have been seen in surveys of low-mass weak-lined
T Tauri stars in regions including $\rho$ Oph, Upper~Sco, Taurus, and
perhaps the Pleiades
\citep{rebull_rotation_2016,david_transient_2017,stauffer_orbiting_2017,stauffer_rotevol_2018,rebull_usco_2018,rebull_rotation_2020}.
These surveys were performed using K2 \citep{howell_k2_2014}.  We
downloaded a few of these lightcurves from MAST, opting for the
EVEREST reductions \citep{luger_everest_2016,luger_update_2018}.  They
are plotted in Figure~\ref{fig:brethren}.

These lightcurves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.
As \citet{stauffer_orbiting_2017} explain,
these  objects are morphologically distinct from ``scallop shell' lightcurves,
and are present in stars at more advanced evolutionary disk stages than
the ``dipper'' stars \citep{ansdell_young_2016,cody_manyfaceted_2018}.
The persistent and transient flux dip stars have the following properties.
\begin{enumerate}
  \item Weak-lined T Tauri stars.
  \item Spectral type typically M2-M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item Ages $\lesssim$ 100$\,$Myr.
  \item Stars typically lack spectroscopic accretion indicators 
  \item Phased lightcurves show shallow, angular dips, usually
    superposed on large-amplitude smooth variability. The latter
    is interpreted as stellar rotation.
  \item Rotation is fast---usually between 0.5 and 2.0 days.
  \item Rarely any infrared excess (never any W4 detection; only a few
    cases with W3 detctions).
  \item Can show multiple dips per cycle ({\it e.g.}, EPIC 203692610,
    RIK-210).
  \item Dip depths, durations, and phases can vary over just a few cycles
    ({\it e.g.}, EPIC 204143627).
  \item Dip depths can change after flares.
  \item Rare at a population level: $\lesssim 1\%$ of relevant stars
    \citep{rebull_usco_2018}.
\end{enumerate}
The 10.74 hour signal in \ptfo\ meets all of these criteria.
This is the first connection of \ptfo\ with this class of objects
because the TESS lightcurve was needed to resolve the different
rotation signals.

There are two crucial additional points concerning the transient flux
dips.  First, the dip durations seem to scale linearly with their
periods \citep[][Figure~26]{stauffer_orbiting_2017}.  In the usual
idealized limits, the transit duration $T$ of a point source across
the stellar disk scales as $T \propto R_\star (P/M_\star)^{1/3}$
\citep{winn_exoplanet_2010}.  While the shortest period
$\approx$0.5-day transient flux dip stars have dip durations
consistent with point sources, at longer periods of 1 to 5 days the
dip durations become many hours, and far too long for a point-source
origin.

Second, between 40-50\% of the transient flux dip stars discovered in
$\rho$~Oph and Upper~Sco show two
Lomb-Scargle periods, and so are apparently binaries
\citep[][Table~1]{stauffer_orbiting_2017}.
This is higher than the corresponding main-sequence companion
fraction of ${\rm CF}_{0.1\text{-}0.5\,M_\odot}^{\rm MS} = 33 \pm 5\%$
\citep{henry_solar_2006,duchene_stellar_2013}.
Low-mass pre-main-sequence stars however have been shown to 
companion fractions up to twice as high in dispersed clusters such as
Upper~Sco and Taurus \citep{kraus_mapping_2008,kraus_mapping_2011}.
A detailed high-resolution imaging survey would be necessary to determine
whether the transient flux dip stars truly have any distinct
population-level binarity properties relative to other young low-mass
stars.

@@

The evidence for binarity in \ptfo\ is as follows.  First, the star is
photometrically twice as bright as stars of the same color in its
kinematic group (Figure~\ref{fig:gaia}.  Second, it shows two distinct
photometric signals.  These points alone suggest binarity
\citep{stauffer_rotevol_2018}.  For the case of \ptfo, there is a
third line of evidence: the Gaia astrometry shows a poor
goodness-of-fit to a single-source model.  While this could be caused
by an additional companion or stellar variability, cluster members
that are equally if not more variable still show \ptfo\ to be an
outlier.  Therefore the astrometric excess is a suggestive third line
of evidence for binarity in \ptfo.  We take these three points
combined to imply that \ptfo\ is probably a binary.

Based on the lack of an infrared excess seen by \citet{yu_tests_2015},
the primordial gas disks seem to be have been depleted\footnote{A
potential confounding factor is completeness. Is Spitzer sensitive to
infrared excesses at the distance of Orion? The series of studies by
\cite{hernandez_spitzer_2006,hernandez_spitzer_ob1_2007,hernandez_spitzer_sig_2007,hernandez_spitzer_2009}
show that it is.} around both stars in \ptfo. The stars are therefore
no longer magnetically locked to their disks.  This is consistent with
the $\approx$half-day periodicities of both rotation signals: young
disked M dwarfs typically rotate with periods of two days or more due
to magnetic locking \citep[{\it e.g.},][]{rebull_rotation_2020}.  If
the two stars are within $\approx$50$\,$AU of each other, as required
by the NIRC2 adaptive optics imaging, then it would also be expected
that the stars would truncate the outer edges of their respective
disks, in a manner seen at the population level in exoplanetary
systems \citep{kraus_impact_2016,moe_impact_2019}.  This hard outer
boundary condition could propagate to the inner disk and affect its
evolution.

The main physical question remaining is clearly what causes the
transient dipping. This is an unsolved problem not only for \ptfo\ but
also for an emerging class of similar young rapidly rotating M-dwarfs.
Some possible explanations are highly disfavored, for reasons that
have been detailed by \citet{david_transient_2017},
\citet{stauffer_orbiting_2017}, and \citet{zhan_complex_2019}.
These disfavored explanations include that the dips are caused by
{\it (i)} eclipsing binaries;
{\it (ii)} ``dipper''-flavor Class-I or Class-II disks;
{\it (iii)} eclipses of prominences;
{\it (iv)} high-latitude accretion hotspots;
{\it (v)} high-latitude starspots;
and
{\it (vi)} dust clouds of reasonable composition.
We also view the possibility of {\it (vii)} tidally disrupted planetary
or cometary material to be implausible, given the synchronicity
between dip and rotation periods, and the stability of signals
observed in some systems.

The explanations that are not yet ruled out include
{\it (i)} transiting clumps of gas at the Keplerian corotation radius;
{\it (ii)} transits of enshrouded protoplanets;
{\it (iii)} occultations of starspots by an optically thick disk.
The first and last explanations offer added appeal because they are
flexible enough to explain not only the transient and persistent-dip
M-dwarfs, but also the ``scallop shell'' M-dwarfs
\citep{stauffer_orbiting_2017}.
Despite this appeal, the possibility of distinct mechanisms 
explaining these
distinct variability classes should remain open.

@@

Given the available evidence,
PTFO$\,$8-8695 seems consistent with the
``transient dipping'' phenomenology observed in many young M dwarfs.
Though it seems rather unlikely to be a planet, understanding \ptfo\
and its analogs will help us understand the natal environments 
of the majority of Earth-sized planets in the Milky Way
\citep{dressing_occurrence_2013}.
